Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-09-25 paper/meg_exps.tex
a

\begin{figure}[t!]
  \begin{minipage}[c]{0.6\textwidth}
    \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_evoked.pdf}
  \end{minipage}\hfill
  \begin{minipage}[c]{0.4\textwidth}
    \caption{Ninety subjects read approximately 2,700 words while their brain activity was recorded with MEG. Top. Average brain response to words (word onset at t=0 ms), as viewed from above the head (red= higher gradient of magnetic flux). Bottom. Each line represents magnetometer, color-coded by spatial position. Posterior responses, typical of primary visual cortex activity, peak around 100 ms after word onset and are followed by an anterior propagation of activity typical of semantic processing in the associative cortices.
    }
    \label{fig:meg_evoked}
  \end{minipage}
\end{figure}

@@

We aim to identify the word features that cause a variation in brain responses. We consider four distinct but colinear features.
%
First, 'Word Length' refers to the total number of letters. Word Length is expected to specifically cause a variation in the early evoked MEG responses (i.e. from 100 ms after stimulus onset) elicited by the retinotopically-tuned visual cortices (e.g. \citep{pegado2014timing}.).
%
Second, 'Word Frequency' indexes how frequently each word appears in Dutch and was derived with the the Zipf logarithmic scale of \citep{van2014subtlex} provided by the WordFreq package \citep{speerwordfreq}. Word Frequency is expected to specifically cause a variation in the late evoked MEG responses (i.e. from 400 ms), because it variably engages semantic processes in the temporal cortices \citep{kutas2011thirty}.
%
Third, 'Word Function' indicates whether each word is a content word (i.e. a noun, a verb, an adjective or an adverb) or a function word (i.e. a preposition, a conjunction, a determinant, a pronoun or a numeral), and was derived from Spacy's part of speech tagger \citep{spacy2}. To our knowledge, this feature has not been thouroughly investigated with MEG. Its causal contribution to reading processes in the brain thus remains unclear.
%
Finally, to verify that B2B and other methods would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of Word Length and Word Frequency:
$dummy = z(length) + z(frequency) + \mathcal{N}$, where $z$ normalizes features and $\mathcal{N}$ is a random vector sampling Gaussian distribution (all terms thus have a zero-mean and a unit-variance).

This procedure yields an $X \in \mathbb{R}^{n \times d_x}$ matrix of $n\approx$ 2700 words by
$d_x=4$ features for each subject. Each of the columns of $X$ is normalized to
have a mean and a standard deviation of 0 and 1 respectively.

@@

We compare B2B to four standard methods: Forward regression, Backward regression, CCA and PLS, as implemented in scikit-learn \citep{sklearn}, and optimized with nested cross validation over twenty $l2$ regularization parameters logarithmically spaced between $10^{-4}$ and $10^4$ (for regression methods) or 1 to 4 canonical components (for cross-decomposition methods).

@@

Each model was implemented for each subject and each time sample independently. Pairwise comparison between models were performed using a Wilcoxon test across subjects (n=90) using the average $\Delta R$ across time.

Corresponding results are shown in Figure~\ref{fig:meg_results}.

@@

Supplementary Figure \ref{fig:meg_supp} shows that the Backward model decodes the dummy variable well above chance level. In addition, the Backward model reveal a similar decoding time course for Word Length and Word Frequency, even though these features are known to specifically influence early and late MEG responses respectively \citep{kutas2011thirty}. These results illustrate that backward modelling cannot be used to estimate the causal contribution of collinear features.

We thus focus on the four remaining methods (i.e. Forward Regression, PLS, CCA, and B2B) and estimate their $\Delta R$ (i.e. the improvement of Y prediction induced by the introduction of a given feature into the model \ref{algorithm:b2b_fi}). Contrary to the Backward Model, none of the models predicted the Dummy Variable to improve the $Y$ prediction: all $\Delta R < 0$ (all $p > .089$).

@@

To assess which model would be most sensitive to these causal discoveries, we compared B2B to other models across subjects (Figure \ref{fig:meg_results} right). For Word Length B2B outperforms all models (all $p < 0.00001$) but CCA ($p=0.0678$). For Word Frequency, B2B outperforms all models (all $p < 0.0006$). For "Word Function", B2B outperforms all models (all $p < 0.0015$). Overall, these results show that B2B reliably outperform standard methods, especially when the effects are difficult to detect.

>>>>

\begin{figure}[t!]
  \begin{minipage}[c]{0.6\textwidth}
    \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_evoked.pdf}
  \end{minipage}\hfill
  \begin{minipage}[c]{0.4\textwidth}
    \caption{Ninety subjects each read approximately 2,700 words while their brain activity was recorded with MEG. Top. Average brain response to words (word onset at t=0 ms), as viewed from above the head (red= higher gradient of magnetic flux). Bottom. Each line represents a magnetometer, color-coded by its spatial position. Posterior responses, typical of primary visual cortex activity, peak around 100 ms after word onset and are followed by an anterior propagation of activity typical of semantic processing in the associative cortices.
    }
    \label{fig:meg_evoked}
  \end{minipage}
\end{figure}

@@

We aim to identify the word features that cause a variation in brain responses. We consider four distinct but colinear features.
%
First, 'Word Length' refers to the total number of letters. Word Length is expected to specifically cause a variation in the early evoked MEG responses (i.e. from 100 ms after stimulus onset) elicited by the retinotopically-tuned visual cortices (e.g. \citep{pegado2014timing}.).
%
Second, 'Word Frequency' indexes how frequently each word appears in Dutch and was derived with the the Zipf logarithmic scale of \citep{van2014subtlex} provided by the WordFreq package \citep{speerwordfreq}. Word Frequency is expected to specifically cause a variation in the late evoked MEG responses (i.e. from 400 ms), because it variably engages semantic processes in the temporal cortices \citep{kutas2011thirty}.
%
Third, 'Word Function' indicates whether each word is a content word (i.e. a noun, a verb, an adjective or an adverb) or a function word (i.e. a preposition, a conjunction, a determinant, a pronoun or a numeral), and was derived from Spacy's part of speech tagger \citep{spacy2}. To our knowledge, this feature has not been thouroughly investigated with MEG. Its causal contribution to reading processes in the brain thus remains unclear.
%
Finally, to verify that B2B and other methods would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of Word Length and Word Frequency:
$dummy = z(length) + z(frequency) + \mathcal{N}$, where $z$ normalizes features and $\mathcal{N}$ is a random vector sampling Gaussian distribution (all terms thus have a zero-mean and a unit-variance). This procedure yields an $X \in \mathbb{R}^{n \times d_x}$ matrix of $n\approx$ 2700 words by
$d_x=4$ features for each subject. Each of the columns of $X$ is normalized to
have a mean and a standard deviation of 0 and 1 respectively.

@@

We compare B2B to four standard methods: Forward regression, Backward regression, CCA and PLS, as implemented in scikit-learn \citep{sklearn}, and optimized with nested cross-validation over twenty $l2$ regularization parameters logarithmically spaced between $10^{-4}$ and $10^4$ (for regression methods) or 1 to 4 canonical components (for cross-decomposition methods).

@@

Each model was implemented for each subject and each time sample independently. Pairwise comparison between models were performed using a two-sided Wilcoxon test across subjects (n=90) using the average $\Delta R$ across time. Corresponding effect sizes are shown in Figure~\ref{fig:meg_results}, and p-values are reported below.

@@

Supplementary Figure \ref{fig:meg_supp} shows that the Backward model decodes the dummy variable well above chance level. In addition, the Backward model reveals a similar decoding time course for Word Length and Word Frequency, even though these features are known to specifically influence early and late MEG responses respectively \citep{kutas2011thirty}. These results illustrate that backward modelling cannot be used to estimate the causal contribution of collinear features.

We thus focus on the four remaining methods (i.e. Forward Regression, PLS, CCA, and B2B) and estimate their $\Delta R$ (i.e. the improvement of Y prediction induced by the introduction of a given feature into the model,as described in Algorithm \ref{algorithm:b2b_fi}). Contrary to the Backward Model, none of the models predicted the Dummy Variable to improve the $Y$ prediction: all $\Delta R < 0$ (all $p > .089$).

@@

To assess which model would be most sensitive to these causal discoveries, we compared B2B to other models across subjects (Figure \ref{fig:meg_results} right). For Word Length B2B outperforms all models (all $p < 0.00001$) but CCA ($p=0.0678$). For Word Frequency, B2B outperforms all models (all $p < 0.0006$). For "Word Function", B2B outperforms all models (all $p < 0.0015$). Overall, these results show that B2B reliably outperforms standard methods, especially when the effects are difficult to detect.

### 2019-09-25 paper/synthetic.tex
a

In most cases, $E$ is not known and AUC can thus not be estimated.

To address this issue, we assess the ability of each model to reliably predict
independent and identically distributed data from $Y$, given all of the $X$
features versus all-but-ones feature $X_{-i}$ (i.e. 'knock-out X'). This procedure
results in two correlation metrics $R_{full}$ and $R_{knockout}$, whose
difference $\Delta R_i = R_{full}-R_{knockout}$ indicates how much each $X_i$
improves the prediction of $Y$.
%beyond what can be predicted from all other features $X_{-i}$.
In our figures, $\Delta R$ is the average of $\Delta R_i$. A higher score means that for prediction, the model relies on individual features rather than combinations of features.

We show in Appendix~\ref{appendix:feature_importance} pseudo-code to assess feature importance for our algorithm as well as baselines. For the Backward Model, feature importance cannot be assessed as there is no prediction.

@@

We show in Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in Appendix) that our method compares favorably to baselines.

>>>>

In most cases, $E$ is not known and AUC can thus not be estimated. To address this issue, we assess the ability of each model to reliably predict
independent and identically distributed data from $Y$, given all of the $X$
features versus all-but-ones feature $X_{-i}$ (i.e. 'knock-out X'). This procedure
results in two correlation metrics $R_{full}$ and $R_{knockout}$, whose
difference $\Delta R_i = R_{full}-R_{knockout}$ indicates how much each $X_i$
improves the prediction of $Y$.
%beyond what can be predicted from all other features $X_{-i}$.
In our figures, $\Delta R$ is the average of $\Delta R_i$. A higher score means that for prediction, the model relies on individual features rather than combinations of features.

We show in Appendix~\ref{appendix:feature_importance} pseudo-code to assess feature importance for our algorithm as well as baselines. For the Backward Model, feature importance cannot be assessed as the $X$ colinearity is never taken into account.

@@

We show in Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in Appendix) that our method outperforms baselines.
